Efficient pedigree recording for fast population genetics simulation

### 2018-10-23 final_format_forwards_paper.tex
final submission version

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{listings}
% \usepackage[T1]{fontenc}

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\justc}{\texttt{C}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\priorityq}[1]{\mbox{\textbf{PriorityQueue}}({#1})}
\newcommand{\listnew}[1]{\mbox{\textbf{List}}(#1)}
\newcommand{\pqpush}[2]{#1.\mbox{\textbf{push}}(#2)}
\newcommand{\pqpop}[1]{#1.\mbox{\textbf{pop}}()}
\newcommand{\listappend}[2]{#1.\mbox{\textbf{append}}(#2)}
\newcommand{\ancsegment}[1]{\mbox{\textbf{Segment}}(#1)}
\newcommand{\taddrow}[2]{#1.\mbox{\textbf{addrow}}(#2)}
\newcommand{\attrparent}[1]{#1.\mbox{parent}}
\newcommand{\attrchild}[1]{#1.\mbox{child}}
\newcommand{\attrleft}[1]{#1.\mbox{left}}
\newcommand{\attrright}[1]{#1.\mbox{right}}
\newcommand{\attrnode}[1]{#1.\mbox{node}}
% mbox to avoid hyphenating
\newcommand{\tsimplify}[1]{\mbox{\textbf{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\textbf{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\textbf{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\textbf{EdgeTable}}(#1)}

@@

%% for submitting with responses to reviews
\newif\ifsubmission
\submissiontrue
% \submissionfalse
\ifsubmission
    % % stuff for submission
    \usepackage[nofiglist,nomarkers,figuresonly]{endfloat}
    \usepackage{setspace}
    % \doublespacing
    \bibliographystyle{plos2015}
    \newcommand\citet{\cite}
    \newcommand\citep{\cite}
    % \renewcommand\includegraphics{\relax}
    % \renewcommand{\includegraphics}[2][]{\relax}
    \renewcommand{\figurename}{Fig.}
    \newcommand{\Figure}{Fig }
    \newcommand{\Figures}{Fig }
\else
    % for the nice version
    \usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
    \bibliographystyle{plainnat}
    \newcommand{\Figure}{{Figure }}
    \newcommand{\Figures}{{Figures }}
\fi
% grrr
\newcommand{\citetx}[2][]{\ifsubmission#1 \cite{#2}\else\citet{#2}\fi}

% Responses to reviews
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\input{review-response-commands}
\newcommand{\responsefile}{review-responses}  % name of the review responses file

\begin{document}

\title{Efficient pedigree recording for fast population genetics simulation}
\author{Jerome Kelleher\footnote{Big Data Institute, University of Oxford},
        Kevin R. Thornton\footnote{Ecology and Evolutionary Biology, University of California at Irvine},
        Jaime Ashander\footnote{Ecology and Evolutionary Biology, University of California at Los Angeles}, and
        Peter L. Ralph\footnote{Institute for Ecology and Evolution, University of Oregon}}
\maketitle

\begin{abstract}
    In this paper we describe how to efficiently record the entire genetic history of a population
    in forwards-time, individual-based population genetics simulations with arbitrary breeding models,
    population structure and demography.
    This approach dramatically reduces the computational burden of tracking individual genomes
    by allowing us to simulate only those loci that may affect reproduction
    (those having non-neutral variants).
    The genetic history of the population is recorded as a succinct tree sequence
    as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    Recording the results of each breeding event requires storage that grows linearly with time,
    but there is a great deal of redundancy in this information.
    We solve this storage problem by providing an algorithm to quickly `simplify'
    a tree sequence by removing this irrelevant history for a given set of genomes.
    By periodically simplifying the history with respect to the extant population,
    we show that the total storage space required is modest
    and overall large efficiency gains can be made over classical forward-time simulations.
    We implement a general-purpose framework for recording and simplifying genealogical data,
    which can be used to make simulations of any population model more efficient.
    We modify two popular forwards-time simulation frameworks to use this new approach
    and observe efficiency gains in large, whole-genome simulations of one to two orders of magnitude.
    In addition to speed, our method for recording pedigrees has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
        can be stored in $O(N \log N + M)$ space,
        making it feasible to store a simulation's entire final generation
        as well as its history.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation 
        of deep history.
    The software for recording and processing tree sequences is named \tskit.
\end{abstract}

% NOTE: I tend to use semantic linebreaks.
% Please excuse the ragged right margins in the source.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Author Summary}

@@

@@

% When tracking neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for these simulations).
Pedigree tracking dramatically reduced runtimes, as shown in Figure~\ref{fig:runtimes_selection},
producing a relative speedup of up to around 50 fold relative to standard simulations that track neutral mutations
(Figure~\ref{fig:relative_speedup_selection}).
Pedigree tracking results in greater relative speedups for larger $N$
and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Figure~\ref{fig:relative_speedup_selection}).
Importantly, runtimes are approximately linear in region size $\rho$ when pedigree tracking
(partially obscured by the log scale of the horizontal axis in Figure~\ref{fig:runtimes_selection}).
%% don't know what an 'edge' is yet
% The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
% each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
In a more limited set of neutral simulations we found the same qualitative behavior,
and a numerically larger speedup by using pedigree tracking (see Appendix~\ref{ss:timing_nosel}).

In our implementation, simulations with pedigree recording
used substantially more RAM than simple forward simulations (see Appendix~\ref{ss:memuse}).
This is unsurprising:
unsimplified tree sequences grow quickly, and so storing history can use arbitrarily much memory.
However, this is not a requirement of the method, only a straightforwards consequence of a speed--memory tradeoff:
the amount of required memory is mostly determined by the interval between simplification steps,
but less frequent simplification reduces overall computation time (see Appendix~\ref{ss:gcinterval}).
In fact, our method could in some situations \emph{reduce} the amount of memory required,
if memory usage in the forwards simulation was dominated by the cost of maintaining neutral genetic variants.

\begin{figure}
    \includegraphics[]{sims/rawspeed_logy}
    \caption{
        \label{fig:runtimes_selection}
    Total run time per single simulation replicate as a function of region length.
    % measured as the scaled recombination parameter $\rho = 4Nr$.
    Line color represents different diploid population sizes ($N$).
    The left figure shows run times for standard simulations including neutral mutations.
    The right column shows run times of simulations that recorded the pedigree
    and added neutral mutations afterwards.
    The dashed line in the right panel shows results for an implementation using \fwdpy{}
    where the pedigree simplification steps were handled in a separate thread of execution
    and fitness calculations were parallelized across four cores.
    Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
    }
\end{figure}

\begin{figure}
    \includegraphics[]{sims/speedup}
    \caption{
        \label{fig:relative_speedup_selection}
        Relative speedup of simulations due to pedigree recording.
        Each line shows the ratio of total run times of standard simulations to
        those of simulations with pedigree recording.
        Data points are taken from Figure~\ref{fig:runtimes_selection}
        for simulations that ran to completion in both cases.
    }
\end{figure}

@@

We now explain what we actually did to achieve this $50\times$ speedup.
The ``pedigree recording'' simulations above recorded information about each new individual
in a collection of tables that together define a \emph{succinct tree sequence}
(or, simply ``tree sequence'').
A {tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
Tree sequences are efficient because branches that are shared by adjacent trees are stored once,
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants is recorded as \emph{sites} and \emph{mutations};
we give an example in Figure~\ref{fig:example_tree_sequence}.
This formulation is derived from the ``coalescence records'' encoding of tree
sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
and generalised to include a more general class of tree topologies.

The \emph{nodes} of a tree sequence
correspond to the vertices in the individual genealogies along the sequence.
Each node refers to a specific, distinct ancestor,
and so has a unique ``time'',
thought of as the node's birth time, which determines the height of any vertices
the node is associated with.
(Note that since each node time is equal to the amount of time since the {birth} of the
corresponding parent, time is measured in clock time, not in meioses.)
The example of Figure~\ref{fig:example_tree_sequence} has five nodes:
nodes 0, 1 and 2 occur at time 0 and are the \emph{samples},
while nodes 3 and 4 represent those ancestors necessary to record their genealogy,
who were born one and two units of time in the past, respectively.

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10.
        The leftmost panels show the tree sequence pictorially in two different ways:
        (top) a sequence of tree topologies; the first tree extends from genomic position 0 to 5,
        and the second from 5 to 10; and
        (bottom) the edges that define these topologies,
        displayed over their corresponding genomic segment
        (for instance, the edge from node 2 to node 4 is present only on the interval from 0 to 5).
        The remaining panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations).
        \label{fig:example_tree_sequence}
    }
\end{figure}

The \emph{edges} define how nodes relate to each other over specific genomic intervals.
Each edge records
% is a tuple $(\ell, r, p, c)$, where
the endpoints $[\ell, r)$ of the half-open genomic interval defining the
spatial extent of the edge;
and the identities $p$ and $c$ of the parent and child nodes
of a single branch that occurs in all trees in this interval.
The spatial extent of the edges defining the topology of Figure~\ref{fig:example_tree_sequence}
are shown in the bottom left panel.
For example, the branch joining nodes 1 to 3 appears in both trees,
and so is recorded as a single edge extending over the whole chromosome.
It is this method of capturing the shared structure between adjacent trees that makes the
tree sequence encoding compact and algorithmically efficient.

@@

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
records two things: its position on the genome and an ancestral state.
For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2.5 with ancestral state `A' and the other at position 7.5 with ancestral state `G'.
If no mutations occur at a given site, all nodes inherit the ancestral state.
Each mutation records three things: the site at which it occurs,
the first node to inherit the mutation, and the derived state.
Thus, all nodes below the mutation's node in the tree will inherit this state,
unless further mutations are encountered.
Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
illustrated by red stars.
The first site, in the left-hand tree,
has a single mutation, which results in node $2$ inheriting the state `T'.
The second site, in the right hand tree, has two mutations:
one occurring over node $3$ changing the state to `C',
and a back mutation over node $1$ changing the state to `G'.

@@

\paragraph{Flexibility.}
To demonstrate the flexibility provided by the Tables API and provide an
implementation that decouples forward simulation internals from transfer of data
to \tskit, we also implemented a version of the simulations described in
``Simulation benchmarks'' separately in Python, described in Appendix~\ref{ss:simupop}.
In this proof-of-concept implementation, the simulation engine (we use \simupop{}, \citet{simupop})
invokes callbacks at critical points of the simulation, and we infer nodes and edges
from the information that is provided. Rows are appended to the tables
one-by-one, and the tables are periodically sorted and simplified to control
memory usage.
Benchmarking results from this implementation are shown (alongside results from \fwdpp{})
for simulations without selection in Appendix~\ref{ss:timing_nosel}:
a relatively modest speedup of around $5 \times$ is achieved, likely due to increased overhead.

>>>>

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage[hidelinks]{hyperref}
\usepackage{nameref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\usepackage{enumitem}
\usepackage{listings}

\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\C}{\mathbb{C}}

\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
%\newcommand{\det}{\mathop{\mbox{det}}}
\newcommand{\supp}{\mathop{\mbox{supp}}}
\newcommand{\sgn}{\mathop{\mbox{sgn}}}
\newcommand{\EE}[1]{\E\!\left[#1\right]}
\newcommand{\PP}[1]{\P\!\left\{#1\right\}}
\newcommand{\PPP}[2]{\P_{#1}\!\left\{#2\right\}}
\newcommand{\EEE}[2]{\E_{#1}\!\left[#2\right]}
\newcommand{\EEsup}[2]{\E^{#1}\!\left[#2\right]}

\newcommand{\bone}{\mathbf{1}}

% These macros are borrowed from TAOCPMAC.tex
\newcommand{\slug}{\hbox{\kern1.5pt\vrule width2.5pt height6pt depth1.5pt\kern1.5pt}}
\def\xskip{\hskip 7pt plus 3pt minus 4pt}
\newdimen\algindent
\newif\ifitempar \itempartrue % normally true unless briefly set false
\def\algindentset#1{\setbox0\hbox{{\bf #1.\kern.25em}}\algindent=\wd0\relax}
\def\algbegin #1 #2{\algindentset{#21}\alg #1 #2} % when steps all have 1 digit
\def\aalgbegin #1 #2{\algindentset{#211}\alg #1 #2} % when 10 or more steps
\def\alg#1(#2). {\medbreak % Usage: \algbegin Algorithm A (algname). This...
  \noindent{\bf#1}({\it#2\/}).\xskip\ignorespaces}
\def\kalgstep#1.{\ifitempar\smallskip\noindent\else\itempartrue
   \hskip-\parindent\fi
   \hbox to\algindent{\bf\hfil #1.\kern.25em}%
   \hangindent=\algindent\hangafter=1\ignorespaces}

\newcommand{\algstep}[3]{\kalgstep #1 [#2] #3 }
\newenvironment{taocpalg}[3]{%
\vspace{1em}%
\algbegin Algorithm #1. ({#2}). #3 }
{\vspace{1em}}

\newcommand{\randomuniform}[0]{\mathcal{R}_U}
\newcommand{\randomdiscrete}[0]{\mathcal{R}_D}
\newcommand{\algref}[1]{#1}

% \newcommand{\tablenotation}[1]{\mathcal{#1}}
% \newcommand{\tableaddrow}[0]{\mbox{add}}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\justc}{\texttt{C}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\priorityq}[1]{\mbox{\textbf{PriorityQueue}}({#1})}
\newcommand{\listnew}[1]{\mbox{\textbf{List}}(#1)}
\newcommand{\pqpush}[2]{#1.\mbox{\textbf{push}}(#2)}
\newcommand{\pqpop}[1]{#1.\mbox{\textbf{pop}}()}
\newcommand{\listappend}[2]{#1.\mbox{\textbf{append}}(#2)}
\newcommand{\ancsegment}[1]{\mbox{\textbf{Segment}}(#1)}
\newcommand{\taddrow}[2]{#1.\mbox{\textbf{addrow}}(#2)}
\newcommand{\attrparent}[1]{#1.\mbox{parent}}
\newcommand{\attrchild}[1]{#1.\mbox{child}}
\newcommand{\attrleft}[1]{#1.\mbox{left}}
\newcommand{\attrright}[1]{#1.\mbox{right}}
\newcommand{\attrnode}[1]{#1.\mbox{node}}
% mbox to avoid hyphenating
\newcommand{\tsimplify}[1]{\mbox{\textbf{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\textbf{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\textbf{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\textbf{EdgeTable}}(#1)}

@@

% % stuff for submission
\usepackage[nofiglist,nomarkers,figuresonly]{endfloat}
\newcommand\citet{\cite}
\newcommand\citep{\cite}
\renewcommand\includegraphics{\relax}
\renewcommand{\includegraphics}[2][]{\relax}
\newcommand{\revpoint}[2]{}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}
% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Efficient pedigree recording for fast population genetics simulation}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
    Jerome Kelleher\textsuperscript{1},
    Kevin R.\ Thornton\textsuperscript{2},
    Jaime Ashander\textsuperscript{3},
    Peter L.\ Ralph\textsuperscript{4,*},
\\
\bigskip
\textbf{1} 
    {Big Data Institute, University of Oxford, UK}
\\
\textbf{2} 
    {Ecology and Evolutionary Biology, University of California at Irvine, USA}
\\
\textbf{3} 
    {Ecology and Evolutionary Biology, University of California at Los Angeles, USA}
\\
\textbf{4} 
    {Institute for Ecology and Evolution, University of Oregon, USA}
\bigskip

% Use the asterisk to denote corresponding authorship and provide email address in note below.
\textbf{*} \url{plr@uoregon.edu}

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
    In this paper we describe how to efficiently record the entire genetic history of a population
    in forwards-time, individual-based population genetics simulations with arbitrary breeding models,
    population structure and demography.
    This approach dramatically reduces the computational burden of tracking individual genomes
    by allowing us to simulate only those loci that may affect reproduction
    (those having non-neutral variants).
    The genetic history of the population is recorded as a succinct tree sequence
    as introduced in the software package msprime,
    on which neutral mutations can be quickly placed afterwards.
    Recording the results of each breeding event requires storage that grows linearly with time,
    but there is a great deal of redundancy in this information.
    We solve this storage problem by providing an algorithm to quickly `simplify'
    a tree sequence by removing this irrelevant history for a given set of genomes.
    By periodically simplifying the history with respect to the extant population,
    we show that the total storage space required is modest
    and overall large efficiency gains can be made over classical forward-time simulations.
    We implement a general-purpose framework for recording and simplifying genealogical data,
    which can be used to make simulations of any population model more efficient.
    We modify two popular forwards-time simulation frameworks to use this new approach
    and observe efficiency gains in large, whole-genome simulations of one to two orders of magnitude.
    In addition to speed, our method for recording pedigrees has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
        can be stored in $O(N \log N + M)$ space,
        making it feasible to store a simulation's entire final generation
        as well as its history.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation 
        of deep history.
    The software for recording and processing tree sequences is named tskit.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author Summary}

@@

\linenumbers

@@

% When tracking neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for these simulations).
Pedigree tracking dramatically reduced runtimes, as shown in Fig~\ref{fig:runtimes_selection},
producing a relative speedup of up to around 50 fold relative to standard simulations that track neutral mutations
(Fig~\ref{fig:relative_speedup_selection}).
Pedigree tracking results in greater relative speedups for larger $N$
and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Fig~\ref{fig:relative_speedup_selection}).
Importantly, runtimes are approximately linear in region size $\rho$ when pedigree tracking
(partially obscured by the log scale of the horizontal axis in Fig~\ref{fig:runtimes_selection}).
%% don't know what an 'edge' is yet
% The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
% each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
In a more limited set of neutral simulations we found the same qualitative behavior,
and a numerically larger speedup by using pedigree tracking (see \nameref{ss:timing_nosel}).

In our implementation, simulations with pedigree recording
used substantially more RAM than simple forward simulations (see \nameref{ss:memuse}).
This is unsurprising:
unsimplified tree sequences grow quickly, and so storing history can use arbitrarily much memory.
However, this is not a requirement of the method, only a straightforwards consequence of a speed--memory tradeoff:
the amount of required memory is mostly determined by the interval between simplification steps,
but less frequent simplification reduces overall computation time (see \nameref{ss:gcinterval}).
In fact, our method could in some situations \emph{reduce} the amount of memory required,
if memory usage in the forwards simulation was dominated by the cost of maintaining neutral genetic variants.

\begin{figure}
    % \includegraphics[]{sims/rawspeed_logy}
    \caption{
        \label{fig:runtimes_selection}
    \textbf{Total run time per single simulation replicate as a function of region length.}
    % measured as the scaled recombination parameter $\rho = 4Nr$.
    Line color represents different diploid population sizes ($N$).
    The left figure shows run times for standard simulations including neutral mutations.
    The right column shows run times of simulations that recorded the pedigree
    and added neutral mutations afterwards.
    The dashed line in the right panel shows results for an implementation using \fwdpy{}
    where the pedigree simplification steps were handled in a separate thread of execution
    and fitness calculations were parallelized across four cores.
    Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
    }
\end{figure}

\begin{figure}
    % \includegraphics[]{sims/speedup}
    \caption{
        \label{fig:relative_speedup_selection}
        \textbf{Relative speedup of simulations due to pedigree recording.}
        Each line shows the ratio of total run times of standard simulations to
        those of simulations with pedigree recording.
        Data points are taken from Fig~\ref{fig:runtimes_selection}
        for simulations that ran to completion in both cases.
    }
\end{figure}

@@

We now explain what we actually did to achieve this $50\times$ speedup.
The ``pedigree recording'' simulations above recorded information about each new individual
in a collection of tables that together define a \emph{succinct tree sequence}
(or, simply ``tree sequence'').
A {tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
Tree sequences are efficient because branches that are shared by adjacent trees are stored once,
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants is recorded as \emph{sites} and \emph{mutations};
we give an example in Fig~\ref{fig:example_tree_sequence}.
This formulation is derived from the ``coalescence records'' encoding of tree
sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
and generalised to include a more general class of tree topologies.

The \emph{nodes} of a tree sequence
correspond to the vertices in the individual genealogies along the sequence.
Each node refers to a specific, distinct ancestor,
and so has a unique ``time'',
thought of as the node's birth time, which determines the height of any vertices
the node is associated with.
(Note that since each node time is equal to the amount of time since the {birth} of the
corresponding parent, time is measured in clock time, not in meioses.)
The example of Fig~\ref{fig:example_tree_sequence} has five nodes:
nodes 0, 1 and 2 occur at time 0 and are the \emph{samples},
while nodes 3 and 4 represent those ancestors necessary to record their genealogy,
who were born one and two units of time in the past, respectively.

\begin{figure}
    \begin{center}
        % \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        \textbf{An example tree sequence with three samples over a chromosome of length 10.}
        The leftmost panels show the tree sequence pictorially in two different ways:
        (top) a sequence of tree topologies; the first tree extends from genomic position 0 to 5,
        and the second from 5 to 10; and
        (bottom) the edges that define these topologies,
        displayed over their corresponding genomic segment
        (for instance, the edge from node 2 to node 4 is present only on the interval from 0 to 5).
        The remaining panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations).
        \label{fig:example_tree_sequence}
    }
\end{figure}

The \emph{edges} define how nodes relate to each other over specific genomic intervals.
Each edge records
% is a tuple $(\ell, r, p, c)$, where
the endpoints $[\ell, r)$ of the half-open genomic interval defining the
spatial extent of the edge;
and the identities $p$ and $c$ of the parent and child nodes
of a single branch that occurs in all trees in this interval.
The spatial extent of the edges defining the topology of Fig~\ref{fig:example_tree_sequence}
are shown in the bottom left panel.
For example, the branch joining nodes 1 to 3 appears in both trees,
and so is recorded as a single edge extending over the whole chromosome.
It is this method of capturing the shared structure between adjacent trees that makes the
tree sequence encoding compact and algorithmically efficient.

@@

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
records two things: its position on the genome and an ancestral state.
For example,
in Fig~\ref{fig:example_tree_sequence} we have two sites, one at position
2.5 with ancestral state `A' and the other at position 7.5 with ancestral state `G'.
If no mutations occur at a given site, all nodes inherit the ancestral state.
Each mutation records three things: the site at which it occurs,
the first node to inherit the mutation, and the derived state.
Thus, all nodes below the mutation's node in the tree will inherit this state,
unless further mutations are encountered.
Three mutations are shown in Fig~\ref{fig:example_tree_sequence},
illustrated by red stars.
The first site, in the left-hand tree,
has a single mutation, which results in node $2$ inheriting the state `T'.
The second site, in the right hand tree, has two mutations:
one occurring over node $3$ changing the state to `C',
and a back mutation over node $1$ changing the state to `G'.

@@

\paragraph{Flexibility.}
To demonstrate the flexibility provided by the Tables API and provide an
implementation that decouples forward simulation internals from transfer of data
to \tskit, we also implemented a version of the simulations described in
``Simulation benchmarks'' separately in Python, described in \nameref{ss:simupop}.
In this proof-of-concept implementation, the simulation engine (we use \simupop{}, \citet{simupop})
invokes callbacks at critical points of the simulation, and we infer nodes and edges
from the information that is provided. Rows are appended to the tables
one-by-one, and the tables are periodically sorted and simplified to control
memory usage.
Benchmarking results from this implementation are shown (alongside results from \fwdpp{})
for simulations without selection in \nameref{ss:timing_nosel}:
a relatively modest speedup of around $5 \times$ is achieved, likely due to increased overhead.
